---
title: "Introduction to R"
---

```{r opts, echo = FALSE}
#suppress the warnings and other messages from showing in the knitted file.
knitr::opts_chunk$set(
  fig.width  = 8, 
  fig.height = 5, 
  fig.path   = 'images/01/',
  echo       = TRUE, 
  warning    = FALSE, 
  message    = FALSE
)
```

## Learning Objectives {.tabset}

### Basic

1. Access [Glasgow Psychology RStudio](http://rstudio1.psy.gla.ac.uk) online
2. Create a simple data table
3. Load data from a CSV and Excel files

### Intermediate

### Advanced


## Prep

* Read [Chapter 1: Introduction](http://r4ds.had.co.nz/introduction.html) in *R for Data Science*


## Resources

* [Chapter 11: Data Import](http://r4ds.had.co.nz/data-import.html) in *R for Data Science*


## Examples

```{r, results = 'hide', warning = FALSE, message = FALSE}

# libraries needed for these examples
library(tidyverse)

```




## Exercises

1. Create a tibble with the name, age, and sex of 3-5 members of your family. 

    What are three ways to look at the data and table setup?

    <div class="solution"><button>Solution</button>
    ```{r}
    # you can do this with an inline csv file using read_csv
    family <- read_csv("name,  age, sex
                        Lisa,   40, female
                        Ben,    41, male
                        Robbie, 10, male")
    
    # or you can do this with the tibble function
    family <- tibble(name = c("Lisa", "Ben", "Robbie"),
                     age = c(40, 41, 10),
                     sex = c("female", "male", "male") )
    
    glimpse(family)
    head(family)
    View(family)
    ```
    </div>

2. Download the dataset [disgust_scores.csv](data/disgust_scores.csv) and read it into a table.

    <div class="solution"><button>Solution</button>
    ```{r}
    disgust <- read_csv("data/disgust_scores.csv")
    ```
    </div>
    
3. What are the different values in the `sex` column?

  <div class="solution"><button>Solution</button>
    ```{r}
    unique(disgust$sex)
    ```
    </div>
    
4. Override the default column specifications to make `sex` a factor and skip the `id` column.

    <div class="solution"><button>Solution</button>
    ```{r}
    my_cols <- cols(
      id = col_skip(),
      user_id = col_integer(),
      sex = col_factor(c("female", "male", "intersex")),
      date = col_datetime(format = ""),
      moral = col_double(),
      pathogen = col_double(),
      sexual = col_double()
    )
    
    disgust <- read_csv("data/disgust_scores.csv", col_types = my_cols)
    head(disgust)
    ```
